Hyperplane Elimination for Quickly Enumerating Local Optima

### 2015-09-10 Hyperplane Enumeration/main.tex
Added text to the experiments section

\section{Mk Landscapes as a tool for Problem Generalization}
An Mk Landscape~\cite{whitley:2015:mk} is any function
who's value is equal to the sum of a set of subfunctions.
Each subfunction uses a small subset of variables from the original function's input.
Combined with limitations on the number of subfunctions and the size of each subfunction's
subset, Mk Landscapes allow for highly efficient local~\cite{whitley:2012:constant,chicano:2014:ball}
and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search.

@@

\section{Improved Enumeration Ordering}
As a final efficiency, the order in which variables in $solution$ are $index$ed can be reordered.
When a move is fitness-improving,
the amount of search space that is skipped depends on how high its lowest $index$ dependency is. Therefore,
by rearranging the order to make its lowest $index$ dependency higher, more search space can be skipped. 
Figure~\ref{fig-enumerate} shows how changing the $index$ order of variables
improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
**000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering that becomes ****0000.
Now whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
skips 4 times as many solutions.

@@

\section{Experiments}
TODO Explain experimental setup details here.

\subsection{Finding Local Optima}
TODO Concatenated Traps using $N>40$ has slope of 2.
\begin{figure*}
  \centering
  \includegraphicsfit{length-method}
  \caption{Comparison of how each method scales with problem size on log-linear scales. Each point is the mean
           runtime over 30 instances and each line the linear model. All confidence intervals too tight to see.}
  \label{fig-length-method}
\end{figure*}

\begin{table}
	\centering
	\caption{Value of $m$ when fitting a model to $y = 2^c2^{mN}$ where $y$ is number of seconds required to
	         find all local optima of a problem using $N$ bits. Optimal refers to the same model where
	         $y$ is the number of local optima.}
	\begin{tabular}{|r|c|c|c|c|}
	  \hline
	    & \textbf{Gray-Box} & \textbf{Hyper} & \textbf{Hyper-Reorder} & \textbf{Optimal} \\ \hline
	  Concatenated Traps & 0.9809 & 0.5765 & 0.1601 & 0.2000 \\ \hline
    Adjacent NKq & 0.9922 & 0.5972 & 0.3411 & 0.3603 \\ \hline
    Random NKq & 0.9920 & 0.6202 & 0.3984 & 0.3531 \\ \hline
    Ising Spin Glass & 0.9821 & 0.6576 & 0.5909 & 0.6015 \\ \hline
    MAX-kSAT & 0.9949 & 0.8893 & 0.7735 & 0.5393 \\ \hline
  \end{tabular}
  \label{table-results}
\end{table}

\begin{figure*}
  \centering
  \includegraphicsfit{boxplot-method}
  \caption{Comparison of completion time variance for the largest size of each problem
           where all three methods were fully successful.}
  \label{fig-boxplot-method}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphicsfit{slope-method}
  \caption{Estimated slope with 95\% confidence intervals for each method to find $r$-bit local optima. Slope is $m$ in the model $y = 2^c2^{mN}$
           where $y$ is number of seconds required to
	         find all local optima of a problem using $N$ bits.}
  \label{fig-boxplot-method}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphicsfit{length-radius}
  \caption{Number of $r$-bit local optima as $N$ increases for each problem.}
  \label{fig-boxplot-method}
\end{figure*}

>>>>

\section{Mk Landscapes as a tool for Problem Generalization}
\label{sec-mk}
An Mk Landscape~\cite{whitley:2015:mk} is any function
who's value is equal to the sum of a set of subfunctions.
Each subfunction uses a small subset of variables from the original function's input.
Combined with limitations on the number of subfunctions and the size of each subfunction's
subset, Mk Landscapes allow for highly efficient local~\cite{whitley:2012:constant,chicano:2014:ball}
and global~\cite{goldman:2015:GBO,tintos:2015:partitioncross} search.

@@

\section{Improved Enumeration Ordering}
\label{sec-reorder}
As a final efficiency, the order in which variables in $solution$ are $index$ed can be reordered.
When a move is fitness-improving,
the amount of search space that is skipped depends on how high its lowest $index$ dependency is. Therefore,
by rearranging the order to make its lowest $index$ dependency higher, more search space can be skipped. 
Figure~\ref{fig-enumerate} shows how changing the $index$ order of variables
improves $m_i$'s lowest $index$ dependency from 2 to 4. Consider that before reordering $m_i$ allows the hyperplane
**000000 to be skipped by Algorithm~\ref{alg-enumerate}, while after reordering that becomes ****0000.
Now whenever $m_i$ is a fitness improvement, Line~\ref{alg-enumerate-bincheck} in Algorithm~\ref{alg-enumerate}
skips 4 times as many solutions.

@@

\section{Experiments}
We compare three methods for finding all local optima: Gray-Box, Hyper, and Hyper-Reorder.
The Gray-Box method refers to previous work~\cite{ochoa:2015:crossovernetworks} which does
not perform hyperplane elimination, with Hyper and Hyper-Reorder being hyperplane elimination
without and with the enumeration order improvements discussed in Section~\ref{sec-reorder}.

For each of the five types of Mk Landscape outlined in Section~\ref{sec-mk}, we tested all
algorithms for all $N$ in $[15..100]$ and $r$ in $[1..4]$. For each problem size we generated
30 problem instances, with each $r$ of each algorithm run once per instance. Variables were randomly
ordered in the genome for problems with natural structure. For example, variables in the same $s_i$
of an Adjacent NKq instance are not likely to be consecutive in the genome. However, $s_i$ still overlaps
$s_{i+1}$ in all but 1 variable.

Each algorithm was given a maximum of four hours on a 2.5GHz Intel Xeon E5-2670v2 processor to find
all $r$-bit local optima. Any run which failed was rerun once to prevent cluster issues from introducing
noise. For the same reason any run which was over ten times slower than the next slowest run of the
same configuration with a different seed was rerun. This resulted in 33 out of 36,000 runs being rerun.
If the algorithm failed to complete in the alloted time for an instance,
we declared it unsuccessful for that combination of $r$ and $N$ of that problem. All results
we report are for successful combinations. For timing purposes local optima were counted but not
recorded.

\subsection{Finding Local Optima}
Our first experiment is to determine the empirical complexity of each method with respect to problem size.
Of the 681 configurations where at least one method was successful, Hyper-Reorder had the lowest
mean time to find all local optima all but 2 times. In both cases Hyper-Reorder was within a tenth of a second
of the best method.
Figure~\ref{fig-boxplot-method} shows the variance in runtime for the largest problem size all three methods
were successful. In all cases it is clear that Hyper-Reorder outperforms the other methods.

\begin{figure*}
  \centering
  \includegraphicsfit{boxplot-method}
  \caption{Comparison of completion time variance for the largest size of each problem
           where all three methods were successful at finding all 1-bit local optima.}
  \label{fig-boxplot-method}
\end{figure*}

More critically, Figure~\ref{fig-length-method} shows how each method scales with problem size.
Not only is Hyper-Reorder out performing the alternatives on all sizes, the rate at which it
scales to problem complexity is better. Table~\ref{table-scaling} shows the model fit for each
method on each problem. This value is the slope of each line shown in Figure~\ref{fig-length-method}.
As expected Gray-Box scales at almost exactly $2^N$. Using Hyperplane elimination without
reordering reduces this complexity, with reordering making further improvements. In the final
column we should the growth rate of the number of local optima. It is impossible for any method
which enumerates all local optima to grow slower than this rate. On Concatenated Traps, Adjacent
NKq, and Ising Spin Glass, Hyper-Reorder has an estimated slope better than optimal. This is
due to lower order terms dominating runtime for small problem sizes, as can be see in
Figure~\ref{fig-length-method}. For instance, fitting
the curve only to problems where $N>60$ makes Hyper-Reorder have $m=0.199$ on Concatenated Traps and
$m=0.358$ on Adjacent NKq.

With the exception of MAX-kSAT, Hyper-Reorder's time required to find all local optima
appears to be growing at the same rate as the total number of local optima. We suspect
some of this deficiency on MAX-kSAT is due to its significantly higher number of non-linear
interactions. Concatenated Traps, Adjacent NKq, and Ising Spin Glass all have exactly 4
non-linear relationships per variable. Random NKq and MAX-kSAT both increase with problem
size, with Random NKq ranging between 5 and 6. With $N=15$, MAX-kSAT has on average nearly
12 non-linear relationships per variable. By $N=30$ that grows to just over 17. Having each
variable depend on over half the genome significantly limits the size of hyperplanes which
can be eliminated.

\begin{figure*}
  \centering
  \includegraphicsfit{length-method}
  \caption{Comparison of how each method scales with problem size when finding 1-bit local optima
           on log-linear scales. Each point is the mean
           runtime over 30 instances and each line the linear model. All confidence intervals too tight to see.}
  \label{fig-length-method}
\end{figure*}

\begin{table}
	\centering
	\caption{Value of $m$ when fitting a model to $y = 2^c2^{mN}$ where $y$ is number of seconds required to
	         find all 1-bit local optima of a problem using $N$ bits. Optimal refers to the same model where
	         $y$ is the number of local optima.}
	\begin{tabular}{|r|c|c|c|c|}
	  \hline
	    & \textbf{Gray-Box} & \textbf{Hyper} & \textbf{Hyper-Reorder} & \textbf{Optimal} \\ \hline
    Concatenated Traps & 0.9809 & 0.5831 & 0.1620 & 0.2000 \\ \hline
    Adjacent NKq & 0.9922 & 0.5972 & 0.3419 & 0.3603 \\ \hline
    Random NKq & 0.9920 & 0.6229 & 0.3989 & 0.3531 \\ \hline
    Ising Spin Glass & 0.9821 & 0.6722 & 0.5976 & 0.6015 \\ \hline
    MAX-kSAT & 0.9949 & 0.8922 & 0.7737 & 0.5393 \\ \hline
  \end{tabular}
  \label{table-scaling}
\end{table}

@@

\begin{figure*}
  \centering
  \includegraphicsfit{slope-method}
  \caption{Estimated slope with 95\% confidence intervals for each method to find $r$-bit local optima. Slope is $m$ in the model $y = 2^c2^{mN}$
           where $y$ is number of seconds required to
	         find all local optima of a problem using $N$ bits.}
  \label{fig-slope-method}
\end{figure*}

@@

\begin{figure*}
  \centering
  \includegraphicsfit{length-radius}
  \caption{Number of $r$-bit local optima as $N$ increases for each problem.}
  \label{fig-length-radius}
\end{figure*}
